Text embedding models yield high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-02-01 paper/main.tex
minor polish/edits

A common approach to assessing students' learning is to compile a set of quiz
questions, tally up the proportion of correctly answered questions, and provide
the student with feedback in the form of a letter grade. Although a numerical
or letter grades give \textit{some} indication about whether the student has
mastered the to-be-learned material, any single measure of performance on a
complex task is at risk of conflating underlying factors, losing relevant
information, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes in detail a student's
knowledge, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion correct scores or letter
grades can be used to instead obtain much more detailed insights into what the
student knows at the time they took the quiz.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overalp
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures 1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time range. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's content. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

We hypothesized that a topic model trained on transcripts of the two lectures,
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices) then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question sets (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = 0.809, p < 0.001$, 95\% confidence interval
(CI)~$= [0.633, 0.962]$), and the average topic weights from lecture 2 are
strongly correlated with the average topic weights from lecture 2 questions
($r(13) = 0.728, p = 0.002, CI = [0.456, 0.920]$). At the same time, the
average topics from two lectures are \textit{negatively} correlated ($r(13) =
-0.634, p = 0.011, CI = [-0.924, -0.237]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between topic vectors for the lectures and each question set is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variance in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001, CI = [0.696, 0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539,
0.958]$). However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video versus lecture 2 questions; lecture 2 video
versus lecture 1 questions) were \textit{negatively} correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondances (at a
coarse scale) between the lectures and (held-out) questions.

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether our text embedding model could be used
to automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors for each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\nameref{subsec:knowledge-maps}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

We developed a computational framework that uses short multiple choice quizzes
to provide nuanced insights into what learners know and how their knowledge
changes with training. First, we show that our approach can automatically match
up the conceptual content of individual questions with the corresponding
moments in lecture videos where those concepts were presented
(Fig.~\ref{fig:question-correlations}). Next, we demonstrate how we can
estimate moment-by-moment ``knowledge traces'' that reflect how much knowledge
participants have before and after watching each lecture video
(Fig.~\ref{fig:knowledge-timeseries}). We also show that these knowledge
estimates can generalize to held-out questions (Fig.~\ref{fig:predictions}).
Finally, we use our framework to construct visual maps that provide snapshot
estimates of how much participants know about any concept within the scope of
our text embedding model, and how much their knowledge changes with training
(Fig.~\ref{fig:knowledge-maps}).

Over the past several years, the global pandemic has forced many educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
downloaded and shared by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized or automated? How can we provide
How and why do learning needs and goals vary across people? How might we lower
barriers to achieving a high quality education?

@@

Beyond specifically predicting what people \textit{know}, the fundamental ideas
we develop here also relate to the notion of ``theory of mind'' of other
individuals~\citep{GoldWinn12, KansEtal15, Melt11}. Considering others' unique
perspectives, prior experiences, knowledge, goals, etc., can help us to more
effectively interact and communicate~\citep{ShaoEtal18, StepBaer06, Ratk18}.
One could imagine future extensions of our work (e.g., analogous to the
knowledge and learning maps shown in Fig.~\ref{fig:knowledge-maps}), that
attempt to characterize how well-aligned different people's knowledge bases or
backgrounds are. In turn, this might be used to model how knowledge flow not
just between teachers and students, but between friends having a conversation,
individuals out on a first date, participants at a business meeting, doctors
and patients, experts and non-experts, political allies and adversaries, and
more. For example, the extent to which two people's knowledge maps ``match'' in
a given region of text embedding space might serve as a predictor of how
effectively the people will communicate about the corresponding conceptual
content.

Ultimately, our work suggests a rich new line of questions about the ``form''
of knowledge, how knowledge changes over time, and how we might map out the
full space of what an individual knows. Our finding that detailed estimates
about knowledge may be obtained from short quizzes shows one way that
traditional approaches to evaluation in education might be extended. We hope
that these advances might help pave the way for new ways of teaching or
delivering educational content that are tailored to individual students'
learning needs and goals.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture 1, the second between
lectures 1 and 2, and the third after viewing lecture 2 (Fig.~\ref{fig:experiment}).
The questions appearing on each quiz, for each participant, were randomly
chosen from the full set of 39, with the constraints that (a) each quiz contain 5
questions about lecture 1, 5 questions about lecture 2, and 3 questions about general physics knowledge, and
(b) each question appear exactly once for each participant. The order of questions on each quiz,
and the order of answer options for each question, were also randomized. Our
experimental protocol was approved by the Committee for the Protection of
Human Subjects at Dartmouth College.  We used the experiment to develop
and test our computational framework for estimating knowledge and learning.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
conceptually related or that tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic. (Unless, otherwise noted, we used $k = 15$
topics.)

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual 
transcriptions of all videos for closed captioning. However, such transcripts 
would not be readily available in all contexts to which our framework could 
potentially be applied. Khan Academy videos are hosted on the YouTube 
platform, which additionally provides automated captions. We opted to use 
these automated transcripts (which, in prior work, we have found are 
sufficiently near-human quality yield reliable data in behavioral studies; 
\citealp{ZimaEtal18}) when developing our framework in order to make it 
more easily extensible and adaptable by others in the future.

@@

Although UMAP is not fully invertible, due to its clustering step, it is
possible to approximately invert the embedded low-dimensional points. In our
application, we sought to embed topic vectors, whose elements are always
non-negative. Because the inversion step is inexact, sometimes it results in
negative-valued vectors (even if the original observations were all
non-negative), which are incompatable with the topic modeling framework. To
protect against this issue, we log-transformed the topic vectors prior to
embedding them in the 2D space. When we inverted the embedded vectors (e.g., to
estimate topic vectors or word clouds, as in Fig.~\ref{fig:knowledge-maps}C),
we passed to inverted (log-transformed) values through the exponential function
to obtain a vector of non-negative values.

After embedding both lectures' topic trajectories and the topic vectors of
every question, we defined a rectangle enclosing the 2D projections of the
lectures' and quizzes' embeddings. We then sampled points from a regular $100
\times 100$ grid of coordinates that evenly tiled this enclosing rectangle. We
sought to estimate participants' knowledge (and learning, i.e., changes in
knowledge) at each of the resulting 10K coordinates.

>>>>

A common approach to assessing students' learning is to compile a set of quiz
questions, tally up the proportion of correctly answered questions, and provide
the student with feedback in the form of a letter grade. Although numerical or
letter grades give \textit{some} indication about whether the student has
mastered the to-be-learned material, any univariate measure of performance on a
complex task is at risk of conflating underlying factors, losing relevant
information, and so on. For example, consider the relative utility of the
imaginary map described above that characterizes in detail a student's
knowledge, versus a single annotation saying that the student answered 85\% of
their quiz questions correctly, or that they received a `B'. Here, we show that
the same quiz data required to compute proportion correct scores or letter
grades can be used to instead obtain much more detailed insights into what the
student knows at the time they took the quiz.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures 1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time range. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's text. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices) then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question sets (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = 0.809, p < 0.001$, 95\% confidence interval
(CI)~$= [0.633, 0.962]$), and the average topic weights from lecture 2 are
strongly correlated with the average topic weights from lecture 2 questions
($r(13) = 0.728, p = 0.002, CI = [0.456, 0.920]$). At the same time, the
average topics from two lectures are \textit{negatively} correlated ($r(13) =
-0.634, p = 0.011, CI = [-0.924, -0.237]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between topic vectors for the lectures and each question set is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001, CI = [0.696, 0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539,
0.958]$). However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video versus lecture 2 questions; lecture 2 video
versus lecture 1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondances (at a
coarse scale) between the lectures and (held-out) questions.

Although a single lecture may be organized around a single broad theme at a
coarse scale, at a finer scale each moment of a lecture typically covers a
narrower range of content. We wondered whether a text embedding model trained
on the lectures' transcripts might capture some of this finer scale content.
For example, if a particular question asks about the content from one small
part of a lecture, we wondered whether the text embeddings could be used to
automatically identify the ``matching'' moment(s) in the lecture. When we
correlated each question's topic vector with the topic vectors from each second
of the lectures, we found some evidence that each question is temporally
specific (Fig.~\ref{fig:question-correlations}). In particular, most questions'
topic vectors were maximally correlated with a well-defined (and relatively
narrow) range of timepoints from their corresponding lectures, and the
correlations fell off sharply outside of that range. We also examined the
best-matching intervals for each question qualitatively by comparing the text
of the question to the text of the most-correlated parts of the lectures.
Despite that the questions were excluded from the text embedding model's
training set, in general we found (through manual inspection) a close
correspondence between the conceptual content that each question covered and
the content covered by the best-matching moments of the lectures. Two
representative examples are shown at the bottom of
Figure~\ref{fig:question-correlations}.

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\nameref{subsec:knowledge-maps}). Next, we sampled points from a $100 \times
100$ grid of coordinates that evenly tiled a rectangle enclosing the 2D
projections of the videos and questions. We used
Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at each of
these 10,000 sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted text embedding
model.

@@

We developed a computational framework that uses short multiple choice quizzes
to provide nuanced insights into what learners know and how their knowledge
changes with training. First, we show that our approach can automatically match
up the conceptual content of individual questions with the corresponding
moments in lecture videos when those concepts were presented
(Fig.~\ref{fig:question-correlations}). Next, we demonstrate how we can
estimate moment-by-moment ``knowledge traces'' that reflect how much knowledge
participants have about each video's content before and after watching each
lecture video (Fig.~\ref{fig:knowledge-timeseries}). We also show that these
knowledge estimates can generalize to held-out questions
(Fig.~\ref{fig:predictions}). Finally, we use our framework to construct visual
maps that provide snapshot estimates of how much participants know about any
concept within the scope of our text embedding model, and how much their
knowledge changes with training (Fig.~\ref{fig:knowledge-maps}).

Over the past several years, the global pandemic has forced many educators to
teach remotely~\citep{MoseEtal21, ShimLee20, KawaEtal21, Whal20}. This change
in world circumstances is happening alongside (and perhaps accelerating)
geometric growth in the availability of high quality online courses on
platforms such as Khan Academy~\citep{Khan04}, Coursera~\citep{Youn12},
EdX~\citep{Kolo13}, and others~\citep{RhoaEtal13}. Continued expansion of the
global internet backbone and improvements in computing hardware have also
facilitated improvements in video streaming, enabling videos to be easily
downloaded and shared by large segments of the world's population. This
exciting time for online course instruction provides an opportunity to
re-evaluate how we, as a global community, educate ourselves and each other.
For example, we can ask: what makes an effective course or training program?
Which aspects of teaching might be optimized or automated? How and why do
learning needs and goals vary across people? How might we lower barriers to
achieving a high quality education?

@@

Beyond specifically predicting what people \textit{know}, the fundamental ideas
we develop here also relate to the notion of ``theory of mind'' of other
individuals~\citep{GoldWinn12, KansEtal15, Melt11}. Considering others' unique
perspectives, prior experiences, knowledge, goals, etc., can help us to more
effectively interact and communicate~\citep{ShaoEtal18, StepBaer06, Ratk18}.
One could imagine future extensions of our work (e.g., analogous to the
knowledge and learning maps shown in Fig.~\ref{fig:knowledge-maps}), that
attempt to characterize how well-aligned different people's knowledge bases or
backgrounds are. In turn, this might be used to model how knowledge (or other
forms of communicable information) flows not just between teachers and
students, but between friends having a conversation, individuals out on a first
date, participants at a business meeting, doctors and patients, experts and
non-experts, political allies or adversaries, and more. For example, the extent
to which two people's knowledge maps ``match'' or ``align'' in a given region
of text embedding space might serve as a predictor of how effectively the
people will communicate about the corresponding conceptual content.

Ultimately, our work suggests a rich new line of questions about the geometric
``form'' of knowledge, how knowledge changes over time, and how we might map
out the full space of what an individual knows. Our finding that detailed
estimates about knowledge may be obtained from short quizzes shows one way that
traditional approaches to evaluation in education might be extended. We hope
that these advances might help pave the way for new ways of teaching or
delivering educational content that are tailored to individual students'
learning needs and goals.

@@

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture 1, the second between
lectures 1 and 2, and the third after viewing lecture 2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture 1, 5 questions about
lecture 2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
semantically related or that tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic. (Unless, otherwise noted, we used $k = 15$
topics.)

As illustrated in Figure~\ref{fig:sliding-windows}A, we start by building up a
corpus of documents using overlapping sliding windows that span each video's
transcript. Khan Academy provides professionally created, manual transcriptions
of all videos for closed captioning. However, such transcripts would not be
readily available in all contexts to which our framework could potentially be
applied. Khan Academy videos are hosted on the YouTube platform, which
additionally provides automated captions. We opted to use these automated
transcripts (which, in prior work, we have found are sufficiently near-human
quality yield reliable data in behavioral studies; \citealp{ZimaEtal18}) when
developing our framework in order to make it more directly extensible and
adaptable by others in the future.

@@

In our application, all of the coordinates we embedded were topic vectors,
whose elements are always non-negative. Although UMAP is an invertible
transformation at the embedding locations of the original data, other locations
in the embedding space will not necessarily follow the same implicit ``rules''
as the original high-dimensional data. For example, inverting an arbitrary
coordinate in the embedding space might result in negative-valued vectors,
which are incompatable with the topic modeling framework. To protect against
this issue, we log-transformed the topic vectors prior to embedding them in the
2D space. When we inverted the embedded vectors (e.g., to estimate topic
vectors or word clouds, as in Fig.~\ref{fig:knowledge-maps}C), we passed the
inverted (log-transformed) values through the exponential function to obtain a
vector of non-negative values.

After embedding both lectures' topic trajectories and the topic vectors of
every question, we defined a rectangle enclosing the 2D projections of the
lectures' and quizzes' embeddings. We then sampled points from a regular $100
\times 100$ grid of coordinates that evenly tiled this enclosing rectangle. We
sought to estimate participants' knowledge (and learning, i.e., changes in
knowledge) at each of the resulting 10,000 coordinates.
